A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2017-02-07 rigid_bodies.tex
intr
introduction

\begin{abstract}
%In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof. Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model. In this paper, we rely on a particular factorization of the rotation matrix to simplify the mechanics of systems containing rigid molecules. Then, we present equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics. We also develop new expressions for computing the virial pressure of rigid-body systems under periodic boundary conditions. Finally, simulations of liquid water are employed to analyze the numerical aspects of the proposed methodology.
\end{abstract}

@@

Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} that requires numerical approximations in order to make possible a microsopic insight into complex physical phenomena. Under those approximations, certain Statistical Mechanics laws might not been necessarily translated into assumptions in Molecular Simulation. For instance, it has already been shown that the unavoidalbe truncation errors associated with the numerical integrations of the equations of motion in Molecular Dynamics (MD) disrupt the equipartition of energy, and the consequence is that the temperature becomes an undefinite property. In a previous paper,\citep{Abreu_2017} we arrived same now regarding the partition of the kinetic energy between translational and rotational degrees of freedom in MD of rigid bodies in the microcanonical (NVE) ensemble. From that observation, we inspected the method introduced by Kamberaj and co-workers\cite{Kamberaj2005}, on which two thermostat chains act separately on the rotational and the translational degrees of freedom and it is not able to reproduced de specified temperature.

The importance of rigid body MD stems from the fact that its employment is not limited to molecules modelled as rigid bodies, like the ubiquitous case of water,  but also it increasingly tool as a coarse-grained strategy for sistems designed as colletions of rigid bodies like molecular motors, nanoparticles, proteins, etc. The method of Kamberaj~\textit{et al.}\cite{Kamberaj2005} was actually used in the simulation of small molecules\cite{Yan_2008,Akimov_2012,Geiger_2013} \cite{Aimoli_2014}, as well as in molecular machines,\cite{Akimov2008, Konyukhov2010} and nanoparticles.\cite{Knorowski2012, Patra2013}

Nos\'{e}-Hoover thermostat is routinely used in Molecular Dynamics to maintain the temperature. It extended phase strategy on which ficticius variables are introduced in order to control the fluctuations of energy and, so the temperature as well. The dynamic of those variables are non-Hamiltonian in nature, which makes the evaluation more difficult, comparing with Hamiltonian dynamics where backward error analysis is done. In fact, it has been used.... In the Nos\'{e}-Hoover thermostat we will carried out a heuristic analysis comparing with SHAKE and Hybrid Monte Carlo (HMC).\cite{Duane1987}  HMC is an exact method in the sense that the truncation erros are not present. And SHAKE is a iterative method that employ lagrange multiplier to solve the holonomic constraints used in MD simulations.For molecules modeled as rigid structures or those containing rigid substructures,\cite{Miller2002} the dynamics of rigid bodies emerges as a suitable approach. This is expected to be computationally more efficient because the constraints are treated implicitly. In addition, and most importantly, it is possible to devise symplectic integrators for simulations in the microcanonical ensemble.

In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble, and introduce the numerical integrators. both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. This differs from the approach of Kamberaj \textit{et al}.,\cite{Kamberaj2005} in which . %We also simplify the calculation of pressure for systems with rigid bodies whose constituent atoms interact via a pairwise potential under periodic boundary conditions. For this, we derive two alternative expressions: %one that depends exclusively on the resultant interaction forces between body pairs and one that corrects the pressure calculated in terms of interactions among freely moving atoms.

>>>>

\begin{abstract}
In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof. Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model. In this paper, we rely on a particular factorization of the rotation matrix to simplify the mechanics of systems containing rigid molecules. Then, we present equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics. We also develop new expressions for computing the virial pressure of rigid-body systems under periodic boundary conditions. Finally, simulations of liquid water are employed to analyze the numerical aspects of the proposed methodology.
\end{abstract}

@@

Molecular Simulation is the computacional realization of Statistical Mechanics\cite{Tuckerman2010} and requires numerical approximations in order to make possible a microsopic insight into complex physical phenomena. Under those approximations, certain Statistical Mechanics laws might not been necessarily traduced into assumptions in Molecular Simulation. For instance, it has already been shown that the unavoidable truncation errors associated with the numerical integrations of the equations of motion in Molecular Dynamics (MD) disrupt the equipartition of energy, and as a consequence the temperature becomes an undefined property\cite{Eastwood_2010}. In a previous paper,\citep{Abreu_2017} we observed the same problem now regarding the partition of the kinetic energy between translational and rotational degrees of freedom in MD of rigid bodies in the microcanonical (NVE) ensemble. That finding brough into question the method introduced by Kamberaj and co-workers\cite{Kamberaj2005}, which couples separate chains of Nos\'{e}-Hoover thermostats on the rotational and translational degrees of freedom and fails to reproduce the specified temperature for certain time step sizes.

The importance of rigid-body Molecular Dynamics stems from the fact that it is not limited to simulate molecules modelled as rigid bodies only, like the ubiquitous case of water,  but also because it is a coarse-grained strategy used for sistems designed as collections of rigid bodies like proteins and nanoparticiles\cite{Knorowski2012, Patra2013}. The approach of Kamberaj~\textit{et al.}\cite{Kamberaj2005}, which is implemented in the LAMMPS software package\cite{Plimpton1995}, has actually been used in the simulation of small molecules\cite{Aimoli_2014}, as well as in molecular motors\cite{Yan_2008,Akimov_2012,Geiger_2013}.

The Nos\'{e}-Hoover thermostat, routinely used in Molecular Dynamics, relies on the ``extended'' phase space on which ficticius variables are introduced in order to control the fluctuations of kinetic energy of the physical variables maintaining the temperature of the physical system in its specified value. The dynamics of those variables introduces non-Hamiltonian components to the equations of motion, which makes its evaluation more limited, comparing with Hamiltonian dynamics where backward error analysis is possible.

In this paper, we focus on the formulation of the equations of motion for rigid bodies in the canonical ensemble, and introduce time reversible and measure-preserving numerical integrators as well. both the translational and rotational degrees of freedom are simultaneously coupled to a unique Nos\'e-Hoover chain (NHC) thermostat. Our starting point to develop the thermostatized equations of motion begining from our previous work on the microcanonical ensemble.

In the Nos\'{e}-Hoover thermostat we will carried out a heuristic analysis comparing with results obtained with Hybrid Monte Carlo (HMC)\cite{Duane1987}, which is an exact method in the sense that the truncation erros are not present, point of reference. We also using SHAKE\cite{Ryckaert1977}, which is an iterative method commonly used in MD simulations, which employes lagrange multiplier to treat the holonomic constraints used in MD simulations. For molecules modeled as rigid structures the dynamics of rigid bodies emerges as a suitable approach. This is expected to be computationally more efficient because the constraints are treated implicitly. The property of being symplectic implies the conservation of volume in phase space. the inverse statement is not true. measure-preserving
